# **Data Exploration**
***

If you have been following along but stopped, we could load our imported data like so:

Add instructions for loading data

***
<details> <summary> If you skipped the data import section click here. </summary>

Add instructions for loading data

</details>
***

## Filter non-informative genes

```{r, eval = FALSE}
keep <- rowSums(assays(sce)$counts > 0) > 10
table(keep)
```

>keep
FALSE  TRUE
23521 30943

```{r, eval = FALSE}
sce_filter <- sce[keep,]
```

## Use perCellQCMetrics to identify low quality or damaged droplets

```{r, eval = FALSE}
qcstats <- scuttle::perCellQCMetrics(sce_filter, subsets = list(Mito = which(rowData(sce_filter)$is_mito)))
```

Warning message about deprecation
> Warning message:
In .per_cell_qc_metrics(assay(x, assay.type), subsets = subsets,  :
  'perCellQCMetrics' is deprecated.
Use 'scrapper::computeRnaQcMetrics' instead.

```{r, eval = FALSE}
lowFeatures <- scuttle::isOutlier(qcstats$detected, type="lower", log=TRUE) #low number of genes
lowLib <- isOutlier(qcstats$sum, type="lower", log=TRUE) #low library depth
highMito <- isOutlier(qcstats$subsets_Mito_percent, type="higher") #high percentage of mitochondrial genes
remove <- lowFeatures | lowLib | highMito
table(remove)
```

>FALSE  TRUE
61046  4943

``{r, eval = FALSE}
filterSuggestions <- list(Features = which(lowFeatures), Library = which(lowLib), Mitochondrial = which(highMito))
ggVennDiagram(filterSuggestions)
```

```{r, echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_0#slide=id.g3e7277bd573_0_0")
```

But since this is using a deprecated function, let's switch to the `scrapper` function that is suggested within the warning message.

However, when I try to use `scrapper::computeRnaQcMetrics`, I get an error that it doesn't work with SummarizedExperiment inputs and to look at the `quickRnaQc.se` function from `scrapper` instead.

```{r, eval = FALSE}
qcmetrics <- scrapper::quickRnaQc.se(sce_filter, subsets = list(Mito = which(rowData(sce_filter)$is_mito)))
table(!qcmetrics$keep)
```

>FALSE  TRUE
61046  4943

We see that this approach is suggesting the removal of the same number of droplets. And from consulting the documentation on the function, it's just a faster/more efficient implementation of the same steps that we did using `scuttle`. This one however, in one line of code, adds `colData` telling us what we should keep in addition to the `sum`, `detected`, and `subset.proportion.Mito` values that the scuttle function previously returned. It also returns the thresholds that it used to from those values to decide on which droplets should be filtered. Previously that required additional `scuttle` functions to do. Both use "adaptive" thresholds that depend upon what is observed in the data and the median absolute deviation that is observed for each metric across all the droplets.

(Add a dropdown section comparing and contrasting this with fixed threshold values as described here: https://bioconductor.org/books/release/scrapbook/quality-control.html#qc-fixed)

```{r, eval=FALSE}
metadata(qcmetrics)$qc$thresholds
```

>$sum
[1] 1089.132

$detected
[1] 917.5423

$subset.proportion
      Mito
0.01257542

```{r, eval = FALSE}
scater::plotColData(qcmetrics, y = "subset.proportion.Mito", colour_by="keep")
qcmetrics$highMito <- qcmetrics$subset.proportion.Mito >= metadata(qcmetrics)$qc$thresholds$subset.proportion[["Mito"]]
plotColData(qcmetrics, y = "subset.proportion.Mito", colour_by = "highMito")
```

```{r, echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_5#slide=id.g3e7277bd573_0_5")
```

```{r, eval = FALSE}
plotColData(qcmetrics, y = "detected", colour_by = "keep")
qcmetrics$lowFeature <- qcmetrics$detected <= metadata(qcmetrics)$qc$thresholds$detected
plotColData(qcmetrics, y = "detected", colour_by = "lowFeature")
```

```{r, echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_12#slide=id.g3e7277bd573_0_12")
```

```{r, eval = FALSE}
plotColData(qcmetrics, y = "detected", colour_by = "lowFeature")
qcmetrics$lowLibrary <- qcmetrics$sum <= metadata(qcmetrics)$qc$thresholds$sum
plotColData(qcmetrics, y = "sum", colour_by = "lowLibrary")
```

```{r, echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_19#slide=id.g3e7277bd573_0_19")
```

What if our filtering were aware of the sample ID information?

```{r, eval = FALSE}
qcmetrics_block <- quickRnaQc.se(sce_filter, subsets = list(Mito = which(rowData(sce_filter)$is_mito)), block = sce_filter$sample_id)
table(!qcmetrics_block$keep)
```

>FALSE  TRUE
61166  4823

We keep a little over 100 droplets we would have otherwise filtered.

```{r, eval = FALSE}
qcmetrics_block$highMito <- qcmetrics_block$subset.proportion.Mito >= metadata(qcmetrics_block)$qc$thresholds$subset.proportion[["Mito"]][qcmetrics_block$sample_id]
qcmetrics_block$lowFeature <- qcmetrics_block$detected <= metadata(qcmetrics_block)$qc$thresholds$detected[qcmetrics_block$sample_id]
qcmetrics_block$lowLibrary <- qcmetrics_block$sum <= metadata(qcmetrics_block)$qc$thresholds$sum[qcmetrics_block$sample_id]
```

```{r, eval = FALSE}
gridExtra::grid.arrange(
    plotColData(qcmetrics_block, y="sum", x="sample_id", colour_by="keep") +
        categoricalHlinesNamed(metadata(qcmetrics_block)$qc$thresholds$sum, levels=NULL, linetype="dashed", color="purple") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        scale_y_log10() +
        ggtitle("Total count"),
    plotColData(qcmetrics_block, y="detected", x="sample_id", colour_by="keep") +
        categoricalHlinesNamed(metadata(qcmetrics_block)$qc$thresholds$detected, levels=NULL, linetype="dashed", color="purple") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        scale_y_log10() +
        ggtitle("Detected features"),
    plotColData(qcmetrics_block, y="subset.proportion.Mito", x="sample_id", colour_by="keep") +
        categoricalHlinesNamed(metadata(qcmetrics_block)$qc$thresholds$subset.proportion[["Mito"]], levels=NULL, linetype="dashed", color="purple") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        ggtitle("Mito proportion"), nrow = 3)
```

```{r, echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_26#slide=id.g3e7277bd573_0_26")
```

And considering the sample violin plots for each of these metrics, we can see that different samples (like 5C and 2E) need more conservative thresholds than other samples. By using less conservative thresholds with the other samples, we end up retaining more data. Because the samples were likely processed separately, this makes sense that they might have different environments driving differences in filtering needs.

```{r, eval = FALSE}
gridExtra::grid.arrange(
    plotColData(qcmetrics_block, y="sum", x="sample_id", colour_by="lowLibrary") +
        categoricalHlinesNamed(metadata(qcmetrics_block)$qc$thresholds$sum, levels=NULL, linetype="dashed", color="purple") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        scale_y_log10() +
        ggtitle("Total count"),
    plotColData(qcmetrics_block, y="detected", x="sample_id", colour_by="lowFeature") +
        categoricalHlinesNamed(metadata(qcmetrics_block)$qc$thresholds$detected, levels=NULL, linetype="dashed", color="purple") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        scale_y_log10() +
        ggtitle("Detected features"),
    plotColData(qcmetrics_block, y="subset.proportion.Mito", x="sample_id", colour_by="highMito") +
        categoricalHlinesNamed(metadata(qcmetrics_block)$qc$thresholds$subset.proportion[["Mito"]], levels=NULL, linetype="dashed", color="purple") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        ggtitle("Mito proportion"), nrow = 3)
```

```{r, echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_32#slide=id.g3e7277bd573_0_32")
```

But is this true? Let's visualize it a couple of ways.

```{r, eval = FALSE}

gridExtra::grid.arrange(
    plotColData(qcmetrics_block, y="sum", x="sample_id", colour_by="lowLibrary") +
        categoricalHlinesNamed(metadata(qcmetrics_block)$qc$thresholds$sum, levels=NULL, linetype="dashed", color="purple") +
        categoricalHlines(unique(qcmetrics_block$sample_id), 1089.132, levels = NULL, linetype = 'dashed', color = "darkgrey") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        scale_y_log10() +
        ggtitle("Total count"),
    plotColData(qcmetrics_block, y="detected", x="sample_id", colour_by="lowFeature") +
        categoricalHlinesNamed(metadata(qcmetrics_block)$qc$thresholds$detected, levels=NULL, linetype="dashed", color="purple") +
        categoricalHlines(unique(qcmetrics_block$sample_id), 917.5423, levels = NULL, linetype = 'dashed', color = "darkgrey") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        scale_y_log10() +
        ggtitle("Detected features"),
    plotColData(qcmetrics_block, y="subset.proportion.Mito", x="sample_id", colour_by="highMito") +
        categoricalHlinesNamed(metadata(qcmetrics_block)$qc$thresholds$subset.proportion[["Mito"]], levels=NULL, linetype="dashed", color="purple") +
        categoricalHlines(unique(qcmetrics_block$sample_id), 0.01257542, levels = NULL, linetype = 'dashed', color = "darkgrey") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        ggtitle("Mito proportion"), nrow = 3)
```

```{r, echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_37#slide=id.g3e7277bd573_0_37")
```

```{r, eval = FALSE}
filterSampleSuggestions <- list(Features = which(qcmetrics_block$lowFeature), Library = which(qcmetrics_block$lowLibrary), Mitochondrial = which(qcmetrics_block$highMito))
ggVennDiagram(filterSampleSuggestions)
```

```{r, echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_42#slide=id.g3e7277bd573_0_42")
```
