# **Data Wrangling**
***

If you have been following along but stopped, we could load our imported data like so:

Add instructions for importing data

***

<details> <summary> If you skipped the data import section click here. </summary>

Add instructions for importing data

</details>
***

Using that zenodo data means there's an overall SingleCellExperiment object for all mice -- we just have to

* handle introns and exons both being represented in the data
* mark mitochondrial genes

Steps in this section

```{r, echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3ea117cd814_0_22#slide=id.g3ea117cd814_0_22")
```

## Handling Introns, Exons, and Mitochondrial genes

```{r, eval = FALSE}
# Extract Exon Ensembl IDs and associated count data
exon_ids <- rowData(snrna_data) %>%
     as.data.frame() %>%
     dplyr::filter(!grepl("-I", gene_id)) %>%
     pull(gene_id)

exon_data <- snrna_data[exon_ids]

# Extract Intron Ensembl IDs and associated count data
intron_ids <- rowData(snrna_data) %>%
     as.data.frame() %>%
     dplyr::filter(grepl("-I", gene_id)) %>%
     pull(gene_id)

intron_data <- snrna_data[intron_ids]

# Simplify Ensembl IDs and find matching IDs between introns and exons

pattern_to_remove <- "\\..*"
replace_string <- ""

## the IDs
exon_ids <- sub(pattern_to_remove, replace_string, exon_ids)
intron_ids <- sub(pattern_to_remove, replace_string, intron_ids)

## the data sources
rownames(exon_data) <- exon_ids
rownames(intron_data) <- intron_ids
rownames(snrna_data) <- sub(pattern_to_remove, replace_string, rownames(snrna_data))

# Subset to only exons and introns with the same Ensembl ID
same_ensembl <- intersect(exon_ids, intron_ids)

# Remove introns and exons with shared IDs from the overall snRNA data
snrna_data_ids_subset <- rownames(snrna_data) %>%
  as.data.frame() %>%
  dplyr::filter(!. %in% same_ensembl) %>%
  pull(.)

# A new SingleCellExperiment object was created, where the sum
# of the UMI counts of spliced mRNA and introns sharing the same
# Ensembl ID was added.
sce <- SingleCellExperiment(assays = list(counts = rbind(counts(snrna_data[snrna_data_ids_subset]),
    counts(exon_data[same_ensembl]) + counts(intron_data[same_ensembl]))))
colData(sce) <- colData(snrna_data)

# Add some metadata columns like condition and replicate number
sce <- sce %>%
  mutate(condition = case_when(str_detect(sample_id, "C") ~ "Control",
                               str_detect(sample_id, "E") ~ "SleepDeprived"
                               )
        ) %>%
  mutate(replicate = factor(str_remove(sample_id, "C|E")))
```

### Annotate the chromosome including mitochondrial genes

```{r, eval = FALSE}
map <- AnnotationDbi::mapIds(EnsDb.Mmusculus.v79, keys = rownames(sce), column = "SEQNAME",
               keytype = "GENEID")
rowData(sce)$CHR <- map
rowData(sce)$is_mito <- if_else(rowData(sce)$CHR == "MT", TRUE, FALSE)
```

> Warning message:
Unable to map 13156 of 54464 requested IDs.

Note that the number of rows matched the length of `map` though.

## Create SingleCellExperiment objects for each mouse

Ambient RNA detection is run on each sample separately, so we want to separate out the samples

```{r, eval = FALSE}
sce_list <- split(sce, sce$sample_id)
names(sce_list) <- make.names(names(sce_list)) #if we don't do this, the names begin with a number and we won't be able to as easily access the different samples
names(sce_list)
```

```
[1] "X1C" "X2E" "X3C" "X4E" "X5C" "X8E"
```

## Ambient RNA detection

Note that Ambient RNA detection should occur before any other filtering (I think even before filtering of non-informative (count 0) genes) because the algorithms rely on being able to model the background distribution of reads.

The classical plot to evaluate whether there may be Ambient RNA in your sample is the Barcode Rank plot that visualizes the total UMI or read counts for each barcode against the barcode's rank (highest to lowest).

Note that though I'm using all of the samples together to make the plot, [according to the OSCA book](https://bioconductor.org/books/3.13/OSCA.advanced/droplet-processing.html#qc-droplets), we want to run ambient RNA detection tools on each sample separately because each sample will have a unique ambient RNA landscape and the tools (at least many of them) are comparing the total counts for each barcode against the total counts for the suspected ambient RNA population to decide what is an "empty" droplet vs a droplet with the data you want.

This approach utilizes base R to find the total counts and ranks (because `DropletUtils` wasn't installing properly for me at first).

```{r, eval = FALSE}
# 1. Calculate total counts (UMIs) per barcode (column)
total_counts <- Matrix::colSums(assays(sce)$counts)
# 2. Filter out barcodes with 0 counts (for speed)
total_counts <- total_counts[total_counts > 0]
# 3. Create a sorted data frame of unique counts and their ranks
bcranks <- data.frame(
Total = sort(total_counts, decreasing = TRUE),
Rank = seq_along(total_counts)
)
# 4. Remove duplicates
bcrank_unique <- bcranks[!duplicated(bcranks$Total), ]
# 5.
bcrank_unique %>%
  ggplot(aes(x=Rank, y = Total)) +
  geom_point() +
  xlab("Barcode Rank") +
  ylab("Total UMI Count") +
  scale_x_log10() +
  scale_y_log10() +
  theme_bw()
```

```{r, echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e545ed5e7a_0_7#slide=id.g3e545ed5e7a_0_7")
```

This approach utilizes `DropletUtils` to compute the sums and barcode ranks

```{r, eval = FALSE}
ranks <- DropletUtils::barcodeRanks(sce)
uniq <- !duplicated(ranks$rank)
ranks_uniq <- ranks[uniq,]
ranks_uniq %>%
  ggplot(aes(x = rank, y = total)) +
  geom_point() +
  xlab("Barcode Rank") +
  ylab("Total UMI Count") +
  scale_x_log10() +
  scale_y_log10() +
  theme_bw()
```

```{r, echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e545ed5e7a_0_64#slide=id.g3e545ed5e7a_0_64")
```

Both plots look the same and the tools that I've used either suggest keeping every droplet, or won't run because it estimates that too much of the sample is contamination. This is the case for `SoupX` which suggested that the estimated contamination level was high (which is to be expected from `snRNA-seq` data): https://github.com/constantAmateur/SoupX/issues/76.

Therefore, I'm turning to a tool that was designed specifically for this task in snRNA-seq data. DIEM or "Debris Identification using Expectation Maximization".

**Working on running this within a script for each sample instead of an interactive session since the EM step takes quite a bit of time.**

```{r, eval = FALSE}
test_sce <- create_SCE(assays(sce_list[["1C"]])$counts)

# Plot total counts ranked
barcode_rank_plot(test_sce)

# DIEM steps
diem_sce_1c <- set_debris_test_set(test_sce)
diem_sce_1c <- filter_genes(diem_sce_1c)
diem_sce_1c <- get_pcs(diem_sce_1c)
diem_sce_1c <- init(diem_sce_1c)
diem_sce_1c <- run_em(diem_sce_1c)
diem_sce_1c <- assign_clusters(diem_sce_1c)
diem_sce_1c <- estimate_dbr_score(diem_sce_1c)

# Evaluate debris scores
sm <- summarize_clusters(diem_sce_1c)
plot_clust(diem_sce_1c, feat_x = "n_genes", feat_y = "score.debris",
           log_x = TRUE, log_y = FALSE)
plot_clust(diem_sce_1c, feat_x = "pct.mt", feat_y = "score.debris",
           log_x = TRUE, log_y = FALSE)

# Call targets using debris score for single-nucleus data
diem_sce_1c <- call_targets(diem_sce_1c, thresh_score = 0.5)

# Call targets by removing droplets in debris cluster(s) for single-cell data
diem_sce_1c <- call_targets(diem_sce_1c, clusters = "debris", thresh = NULL)

```

Likely will provide filtering output suggestions from DIEM (if there are any) to learners using a cooking show approach rather than having them run it themselves.

## Filter non-informative genes

```{r, eval = FALSE}
keep_informative <- rowSums(assays(sce)$counts > 0) > 10
table(keep_informative)
```

```
FALSE  TRUE
23521 30943
```

What thresholds is appropriate here? This code is currently finding genes that have a non-zero count in at least 11 droplets. [As described in this resource](https://link.springer.com/protocol/10.1007/978-1-0716-1307-8_19#Sec23), we want to make sure we're not potentially removing the ability to detect rare cell types and we might be able to use clustering to help tune this number

:::{.column-margin}
And when is the right time to do this step filtering out non-informative genes? I'm fairly certain after ambient RNA detection and before normalization I would expect, but I'm not sure if it needs to be before other QC steps as well.
:::

Note that my computational environment struggles unless I constantly overwrite sce. To increase feasibility, I think

* we need to show learners how to remove large things we won't need again from the environment
* avoid writing intermediates (like `sce_filter`)


## Use perCellQCMetrics to identify low quality or damaged droplets

```{r, eval = FALSE}
qcmetrics <- scuttle::perCellQCMetrics(sce, subsets = list(Mito = which(rowData(sce)$is_mito)))
```

Warning message about deprecation

> Warning message:
In .per_cell_qc_metrics(assay(x, assay.type), subsets = subsets,  :
  'perCellQCMetrics' is deprecated.
Use 'scrapper::computeRnaQcMetrics' instead.

```{r, eval = FALSE}
lowFeatures <- scuttle::isOutlier(qcmetrics$detected, type="lower", log=TRUE) #low number of genes
lowLib <- isOutlier(qcmetrics$sum, type="lower", log=TRUE) #low library depth
highMito <- isOutlier(qcmetrics$subsets_Mito_percent, type="higher") #high percentage of mitochondrial genes
remove <- lowFeatures | lowLib | highMito
table(remove)
```

```
FALSE  TRUE
61051  4938
```

```{r, eval = FALSE}
filterSuggestions <- list(Features = which(lowFeatures), Library = which(lowLib), Mitochondrial = which(highMito))
ggVennDiagram(filterSuggestions)
```

```{r, echo = FALSE, warning = FALSE, out.width=800}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_0#slide=id.g3e7277bd573_0_0")
```

(Add a dropdown section describing MAD (median absolute deviation) and how that is computed as the median of the absolute value of the differences of the observed values from the median of the observed values) and how they're using 3 of those to try to get the outliers from normally distributed data.... plot your data and if it doesn't look normal this approach might not be for you.)

But since this is using a deprecated function, let's switch to the `scrapper` function that is suggested within the warning message.

However, when I try to use `scrapper::computeRnaQcMetrics`, I get an error that it doesn't work with SummarizedExperiment inputs and to look at the `quickRnaQc.se` function from `scrapper` instead.

Note: I'm not necessarily planning to show the whole evolution of planning to use `perCellQCMetrics` and pivoting over to using `quickRnaQc.se`, but am documenting it here since it's a departure from the original plan.

```{r, eval = FALSE}
sce <- scrapper::quickRnaQc.se(sce, subsets = list(Mito = which(rowData(sce)$is_mito)))
table(!sce$keep)
```

```
FALSE  TRUE
61051  4938
```

We see that this approach is suggesting the removal of the same number of droplets. And from consulting the documentation on the function, it's just a faster/more efficient implementation of the same steps that we did using `scuttle`. Both use "adaptive" thresholds that depend upon what is observed in the data and the median absolute deviation that is observed for each metric across all the droplets. This one however has a few differences,

* in one line of code, adds `colData` telling us what we should keep in addition to the `sum`, `detected`, and `subset.proportion.Mito` values that the scuttle function previously returned. Previouly that required additional `scuttle` functions to do.
* It also returns the thresholds that it used to from those values to decide on which droplets should be filtered.
* This one returns a `SingleCellExperiment` object while scuttle only returned a dataframe.

(Add a dropdown section comparing and contrasting this adaptive threshold/MAD approach with fixed threshold values as described here: https://bioconductor.org/books/release/scrapbook/quality-control.html#qc-fixed)

```{r, eval=FALSE}
metadata(sce)$qc$thresholds
```

```
$sum
[1] 1089.368

$detected
[1] 917.9337

$subset.proportion
      Mito
0.01257568
```

```{r, eval = FALSE}
scater::plotColData(sce, y = "subset.proportion.Mito", colour_by="keep")
sce$highMito <- sce$subset.proportion.Mito >= metadata(sce)$qc$thresholds$subset.proportion[["Mito"]]
plotColData(sce, y = "subset.proportion.Mito", colour_by = "highMito")
```

```{r, echo = FALSE, warning = FALSE, out.width=800}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_5#slide=id.g3e7277bd573_0_5")
```

```{r, eval = FALSE}
plotColData(sce, y = "detected", colour_by = "keep")
sce$lowFeature <- sce$detected <= metadata(sce)$qc$thresholds$detected
plotColData(sce, y = "detected", colour_by = "lowFeature")
```

```{r, echo = FALSE, warning = FALSE, out.width=800}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_12#slide=id.g3e7277bd573_0_12")
```

```{r, eval = FALSE}
plotColData(sce, y = "sum", colour_by = "keep")
sce$lowLibrary <- sce$sum <= metadata(sce)$qc$thresholds$sum
plotColData(sce, y = "sum", colour_by = "lowLibrary")
```

```{r, echo = FALSE, warning = FALSE, out.width=800}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_19#slide=id.g3e7277bd573_0_19")
```

What if our filtering were aware of the sample ID information?

Let's save the old output (so if we want to compare), by renaming the column names

```{r, eval = FALSE}
colData(sce) <- colData(sce) %>%
  as.data.frame() %>%
  rename_with( ~ paste0("nb_", .x), #nb for non-block
              .cols = -all_of(c("whitelist", "sample_id", "condition", "replicate"))) %>%
  DataFrame()

metadata(sce)$nb_qc$thresholds <- metadata(sce)$qc$thresholds
```

```{r, eval = FALSE}
sce <- quickRnaQc.se(sce, subsets = list(Mito = which(rowData(sce)$is_mito)), block = sce$sample_id)
table(!sce$keep)
```

```
FALSE  TRUE
61168  4821
```

We keep a little over 100 droplets we would have otherwise filtered.

```{r, eval = FALSE}
sce$highMito <- sce$subset.proportion.Mito >= metadata(sce)$qc$thresholds$subset.proportion[["Mito"]][sce$sample_id]
sce$lowFeature <- sce$detected <= metadata(sce)$qc$thresholds$detected[sce$sample_id]
sce$lowLibrary <- sce$sum <= metadata(sce)$qc$thresholds$sum[sce$sample_id]
```

```{r, eval = FALSE}
gridExtra::grid.arrange(
    plotColData(sce, y="sum", x="sample_id", colour_by="keep") +
        categoricalHlinesNamed(metadata(sce)$qc$thresholds$sum, levels=NULL, linetype="dashed", color="purple") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        scale_y_log10() +
        ggtitle("Total count"),
    plotColData(sce, y="detected", x="sample_id", colour_by="keep") +
        categoricalHlinesNamed(metadata(sce)$qc$thresholds$detected, levels=NULL, linetype="dashed", color="purple") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        scale_y_log10() +
        ggtitle("Detected features"),
    plotColData(sce, y="subset.proportion.Mito", x="sample_id", colour_by="keep") +
        categoricalHlinesNamed(metadata(sce)$qc$thresholds$subset.proportion[["Mito"]], levels=NULL, linetype="dashed", color="purple") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        ggtitle("Mito proportion"),
    nrow = 3)
```

```{r, echo = FALSE, warning = FALSE, out.width=800}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_26#slide=id.g3e7277bd573_0_26")
```

And considering the sample violin plots for each of these metrics, we can see that different samples (like 5C and 2E) need more conservative thresholds than other samples. By using less conservative thresholds with the other samples, we end up retaining more data. Because the samples were likely processed separately, this makes sense that they might have different environments driving differences in filtering needs.

```{r, eval = FALSE}
gridExtra::grid.arrange(
    plotColData(sce, y="sum", x="sample_id", colour_by="lowLibrary") +
        categoricalHlinesNamed(metadata(sce)$qc$thresholds$sum, levels=NULL, linetype="dashed", color="purple") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        scale_y_log10() +
        ggtitle("Total count"),
    plotColData(sce, y="detected", x="sample_id", colour_by="lowFeature") +
        categoricalHlinesNamed(metadata(sce)$qc$thresholds$detected, levels=NULL, linetype="dashed", color="purple") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        scale_y_log10() +
        ggtitle("Detected features"),
    plotColData(sce, y="subset.proportion.Mito", x="sample_id", colour_by="highMito") +
        categoricalHlinesNamed(metadata(sce)$qc$thresholds$subset.proportion[["Mito"]], levels=NULL, linetype="dashed", color="purple") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        ggtitle("Mito proportion"),
    nrow = 3)
```

```{r, echo = FALSE, warning = FALSE, out.width=800}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_32#slide=id.g3e7277bd573_0_32")
```

But is this true? Let's visualize it a couple of ways.

```{r, eval = FALSE}

gridExtra::grid.arrange(
    plotColData(sce, y="sum", x="sample_id", colour_by="lowLibrary") +
        categoricalHlinesNamed(metadata(sce)$qc$thresholds$sum, levels=NULL, linetype="dashed", color="purple") +
        categoricalHlines(unique(sce$sample_id), metadata(sce)$nb_qc$thresholds$sum, levels = NULL, linetype = 'dashed', color = "darkgrey") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        scale_y_log10() +
        ggtitle("Total count"),
    plotColData(sce, y="detected", x="sample_id", colour_by="lowFeature") +
        categoricalHlinesNamed(metadata(sce)$qc$thresholds$detected, levels=NULL, linetype="dashed", color="purple") +
        categoricalHlines(unique(sce$sample_id), metadata(sce)$nb_qc$thresholds$detected, levels = NULL, linetype = 'dashed', color = "darkgrey") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        scale_y_log10() +
        ggtitle("Detected features"),
    plotColData(sce, y="subset.proportion.Mito", x="sample_id", colour_by="highMito") +
        categoricalHlinesNamed(metadata(sce)$qc$thresholds$subset.proportion[["Mito"]], levels=NULL, linetype="dashed", color="purple") +
        categoricalHlines(unique(sce$sample_id), metadata(sce)$nb_qc$thresholds$subset.proportion[["Mito"]], levels = NULL, linetype = 'dashed', color = "darkgrey") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        ggtitle("Mito proportion"),
    nrow = 3)
```

```{r, echo = FALSE, warning = FALSE, out.width=800}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_37#slide=id.g3e7277bd573_0_37")
```

```{r, eval = FALSE}
filterSampleSuggestions <- list(Features = which(sce$lowFeature), Library = which(sce$lowLibrary), Mitochondrial = which(sce$highMito))
ggVennDiagram(filterSampleSuggestions)
```

```{r, echo = FALSE, warning = FALSE, out.width=800}
ottrpal::include_slide("https://docs.google.com/presentation/d/1NwTGnca5G7c_dQLgNpYEgMJp2tWpyjjG7sl4t_vxTPg/edit?slide=id.g3e7277bd573_0_42#slide=id.g3e7277bd573_0_42")
```

note that the [advanced section of the OSCA book](https://bioconductor.org/books/3.14/OSCA.advanced/single-nuclei-rna-seq-processing.html#quality-control-for-stripped-nuclei) suggests that this adaptive thresholding approach may not work for single nuclear data, specifically with respect to mitochondrial reads, because the MAD may be too close to zero and then too much is getting filtered out. Another important note is [mentioned here in this paper](https://link.springer.com/protocol/10.1007/978-1-0716-1307-8_19#Sec23) that certain types of data are going to have much higher mitochondrial activity and that should be taken into account when setting a threshold

:::{.column-margin}
When working through single cell data with Galaxy tutorials, and working with spatial data for another case study, I've seen two sided filters for the number of genes/features and library size where it's not just excluding droplets with a certain value below a minimum, but also excluding those with a certain value above a maximum. However, I haven't seen any OSCA or OSCA related tutorials doing a similar thing. The authors in their vignette also only show a lower test. Should I be utilizing a maximum filter too?
:::

## Applying the filters ....

```{r, eval = FALSE}
sce <- sce[keep_informative,] #non-informative gene filter
```

:::{.column-margin}
should I wait to apply all of the filters at the same time prior to normalization or apply them after each step? scDblFinder maintainer suggests I should https://github.com/plger/scDblFinder/issues/40
:::

## Remove doublets



## Transformation

### Option A: logNormCounts

### Option B: account for nuclei size

Deconvolution

:::{.column-margin}
Since quick clustering is being used within this approach to group cells of similar expression profiles, is it ever recommended to perform cell type assignment prior to deconvolution normalization?
:::


***
