# **Data Import**
***

## Read in data

!<--
When downloading the data with `zen4R` from zenodo, I got warning messages about file integrity. Had to increase the timeout to 600
-->


```{r, eval = FALSE}
zenodo_doi <- "10.5281/zenodo.10091236"
target_directory <- "/fh/fast/leek_j/user/kisaac/ocs_sc/zenodo_files"
zen4R::download_zenodo(doi = zenodo_doi, path = target_directory, timeout = 600)
snrna_data <- readRDS(here::here("zenodo_files/sce_mouse_sleep_snrnaseq_complete.rds"))
```

## Create annotations and SingleCellExperiment object

Using that zenodo data means that the annotations part was already handled and there's an overall SingleCellExperiment object for all mice

## Handling Introns, Exons, and Mitochondrial genes

```{r, eval = FALSE}
# Extract Exon Ensembl IDs and associated count data
exon_ids <- rowData(snrna_data) %>%
     as.data.frame() %>%
     dplyr::filter(!grepl("-I", gene_id)) %>%
     pull(gene_id)

exon_data <- snrna_data[exon_ids]

# Extract Intron Ensembl IDs and associated count data
intron_ids <- rowData(snrna_data) %>%
     as.data.frame() %>%
     dplyr::filter(grepl("-I", gene_id)) %>%
     pull(gene_id)

intron_data <- snrna_data[intron_ids]

# Simplify Ensembl IDs and find matching IDs between introns and exons

pattern_to_remove <- "\\..*"
replace_string <- ""

## the IDs
exon_ids <- sub(pattern_to_remove, replace_string, exon_ids)
intron_ids <- sub(pattern_to_remove, replace_string, intron_ids)

## the data sources
rownames(exon_data) <- sub(pattern_to_remove, replace_string, rownames(exon_data))
rownames(intron_data) <- sub(pattern_to_remove, replace_string, rownames(intron_data))
rownames(snrna_data) <- sub(pattern_to_remove, replace_string, rownames(snrna_data))

# Subset to only exons and introns with the same Ensembl ID
same_ensembl <- intersect(exon_ids, intron_ids)

# Remove introns and exons with shared IDs from the overall snRNA data
snrna_data_ids_subset <- rownames(snrna_data) %>%
  as.data.frame() %>%
  dplyr::filter(!. %in% same_ensembl) %>%
  pull(.)

# A new SingleCellExperiment object was created, where the sum
# of the UMI counts of spliced mRNA and introns sharing the same
# Ensembl ID was added.
sce <- SingleCellExperiment(assays = list(counts = rbind(counts(snrna_data[snrna_data_ids_subset]),
    counts(exon_data[same_ensembl]) + counts(intron_data[same_ensembl]))))
colData(sce) <- colData(snrna_data)

# Add some metadata columns like condition and replicate number
sce <- sce %>%
  mutate(condition = case_when(str_detect(sample_id, "C") ~ "HC", #HC for Home Cage
                               str_detect(sample_id, "E") ~ "SD" #SD for Sleep Deprived
                               )
        ) %>%
  mutate(replicate = factor(str_remove(sample_id, "[A-Za-z]")))
```

### Annotate the chromosome including mitochondrial genes

```{r, eval = FALSE}
map <- AnnotationDbi::mapIds(EnsDb.Mmusculus.v79, keys = rownames(sce), column = "SEQNAME",
               keytype = "GENEID")
rowData(sce)$CHR <- map
rowData(sce)$is_mito <- if_else(rowData(sce)$CHR == "MT", TRUE, FALSE)
```

> Warning message:
Unable to map 13156 of 54464 requested IDs.

Note that the number of rows matched the length of `map` though.

## Create SingleCellExperiment objects for each mouse

```{r, eval = FALSE}
sce_list <- split(sce, sce$sample_id)
```

## Import data as SingleCellExperiment objects



Add instructions for saving data

***
