# **Data Analysis**
***

If you have been following along but stopped, we could load our wrangled data like so:

Add instructions for loading data

***
<details> <summary> If you skipped the data import section click here. </summary>

Add instructions for loading data

</details>
***

<!--
:::{.think_question}
::::{.think_question_header}
Question opportunity
::::
::::{.inner_block}
Pose a question about the data analysis.
<details><summary>Answer</summary>
You can use a collapsible details section to provide an answer
</details>
::::
:::

::: {.column-margin}
We might use a column margin note to add an equation or some other annotation for learners or educators.
:::
-->

## Put the process together with `pipeComp`

We'll want to build at least one function with parameters that control its behavior so that we can use `pipeComp`.

Will build the function in the narrative

```{r, eval = FALSE}

#' A function to drive the normalization and aggregation process
#'
#' @parameter load_data either TRUE or FALSE. Will load data if TRUE, will run the normalization process if FALSE
#' @parameter norm_method either logcounts or deconvolution
#' @parameter aggregation_method either sum or average
#' @parameter aggregation_across either cluster or sample
#' @parameter sce default is NULL, provide if load_data is FALSE. Should be a SingleCellExperiment object that is filtered but not yet normalized. Should have colData columns `sum` and `sample_id`
#'
#' @return aggregated a normalized and aggregated summary of the data


norm_aggregate <- function(norm_method, load_data, aggregation_method, aggregation_across, sce = NULL){

  # check parameter inputs
  stopifnot(load_data %in% c(TRUE, FALSE))
  stopifnot(norm_method %in% c('logcounts', 'deconvolution'))
  stopifnot(aggregation_method %in% c('sum', 'average'))
  stopifnot(aggregation_across %in% c('cluster', 'sample'))

  # normalization

  if (load_data){
    sce <- readRDS("data/wrangled/sce_norm.rds")
  } else {
    if (is.null(sce)){
      stop("Must provide an sce object with filtered, unnormalized data and colData columns `sum` and `sample_id`")
    }
    if (norm_method == "logcounts"){
      sce <- scrapper::normalizeRnaCounts.se(sce,
                                           size.factors = sce$sum,
                                           block = sce$sample_id)
    }
    if (norm_method == "deconvolution"){
      sce$quickClusters <- scran::quickCluster(sce, block = sce$sample_id)
      deconv_sf <- scran::calculateSumFactors(sce, clusters = clusters.block)
      sce <- normalizeRnaCounts.se(sce,
                                 size.factors = deconv_sf,
                                 block = sce$sample_id,
                                 output.name = "deconvcounts",
                                 factor.name = "deconvSizeFactor")
    }
  }

  # aggregation

  if(norm_method == "logcounts"){
    assay_type <- "logcounts"
  } else{
    assay_type <- "deconvcounts"
  }

  if(aggregation_across == "cluster"){
    coi <- "quickClusters"
    if (!coi %in% names(colData(sce))){
      sce$quickClusters <- scran::quickCluster(sce, block = sce$sample_id)
    }
  } else{
    coi <- "sample_id"
  }


  if (aggregation_method == "sum"){
    aggregated <- scrapper::aggregateAcrossCells.se(sce,
                      colData(sce)[[coi]],
                      assay.type = assay_type)
  }

  if (aggregation_method == "average"){
    aggregated <- scrapper::aggregateAcrossCells.se(sce,
                      colData(sce)[[coi]],
                      assay.type = assay_type)
    assay(aggregated, "averages") <- t(t(assays(aggregated)$sums)/aggregated$counts)
  }

  return (aggregated)
}

#make a function to find the stats below

```


### For each normalization option and aggregation combination, working within the quick clusters....

TODO: show how to use pipeComp for these questions....

#### What is the highest expressed gene for each cluster? (What's its expression?)

```{r, eval = FALSE}

result <- apply(assays(agg_cluster_sum)$sums, 2, function(col) {
    c(gene = rownames(agg_cluster_sum)[which.max(col)], max_value = max(col))
})

result %>% t() %>% as.data.frame() %>% mutate(cluster = as.factor(row_number())) %>% left_join(., colData(sce)[c("condition", "quickClusters")] %>% as.data.frame(), by = c("cluster" = "quickClusters") ) %>% unique()
```

#### How many genes have no observed expression within that cluster?

```{r, eval = FALSE}
colSums(assays(agg_cluster_sum)$sums == 0)
```

#### What's the median gene expression value for each cluster? (What gene(s) is it connected to?)

```{r, eval = FALSE}
result <- apply(assays(agg_cluster_sum)$sums, 2, function(col) {
  # Calculate the standard median
  med_val <- round(median(col),2)

  # Find the row index of the value closest to the median
  gene <- rownames(agg_cluster_sum)[which.min(abs(col - med_val))]

  c(gene = gene, median_value = med_val)
})

result %>% t() %>% as.data.frame() %>% mutate(cluster = as.factor(row_number())) %>% left_join(., colData(sce)[c("condition", "quickClusters")] %>% as.data.frame(), by = c("cluster" = "quickClusters") ) %>% unique()
```

***
